---
title: factor model
author: Jialing Deng
date: Mar 23
output:
  pdf_document:
    toc: true
    toc_depth: 1
---

_Note:_ To help show what your homework should look like, these solutions don't
display the code used.  You should look at the Rmd file as well to see those.
(The comments to the code include some notes on why it's done the way it is, and
alternatives.)

```{r, include=FALSE}
# Standard set-up
library(knitr)
knitr::opts_chunk$set(warning=FALSE, message=FALSE)


# Set knitr options for knitting code into the report:
# - Save results so that code blocks aren't re-run unless code changes (cache),
  # _or_ a relevant earlier code block changed (autodep)
  # but don't re-run if only the comments changed (cache.comments)
# - Don't clutter R output with messages or warnings (message, warning)
  # This _will_ leave error messages showing up in the knitted report
# - By default, don't print the code (echo)
opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE, echo=FALSE)
```






```{r}
load("stockData.RData")

```








# Q1

## a. 

```{r}

date <- as.Date(sort(rownames(close_price)))
price_matrix <- as.matrix(close_price)

ymin <- min(price_matrix, na.rm = TRUE)
ymax <- max(price_matrix, na.rm = TRUE)


plot(date, price_matrix[, 1], type = "l",
     xlab = "Date", ylab = "Closing Price",
     main = "Closing Prices of 28 Large US Corporations",
     ylim = c(ymin, ymax))


colors <- rainbow(ncol(price_matrix))
for (i in 1:ncol(price_matrix)) {
  lines(date, price_matrix[, i], col = colors[i])
}




```

## b.
```{r}
price_matrix <- as.matrix(close_price)
log_returns_matrix <- diff(log(price_matrix)) 
log_returns <- as.data.frame(log_returns_matrix)
log_return_dates <- as.Date(rownames(close_price))[-1]
plot(log_return_dates, log_returns_matrix[,1], type = "n",
     xlab = "Date", ylab = "Log Daily Return",
     main = "Log Daily Returns of 28 Large US Corporations")


for (i in 1:ncol(log_returns_matrix)) {
  lines(log_return_dates, log_returns_matrix[, i],col = colors[i])
}


```
Yes, the log returns appear more comparable than the raw closing prices, as the 
log returns are centered around zero, with similar fluctuations for most stocks.


\newpage

### Q2.

### a.
```{r}
ba_log_returns <- log_returns$BA  
hist(ba_log_returns, breaks = 30, probability = TRUE, 
     col = "skyblue", 
     main = "Histogram of Boeing (BA) Log Returns",
     xlab = "Log Return", ylab = "Density")
```
### b. 
```{r}


mu_mle <- mean(ba_log_returns)
sigma_mle <- sd(ba_log_returns)

hist(ba_log_returns, breaks = 30, probability = TRUE,
     col = "skyblue",
     main = "Histogram of Boeing (BA) Log Returns",
     xlab = "Log Return", ylab = "Density")

curve(dnorm(x, mean = mu_mle, sd = sigma_mle), 
      col = "red", lwd = 2, add = TRUE)

```
The Gaussian distribution appears to fit the histogram of Boeing log returns well in 
the center, capturing the general bell-shaped curve around the mean.However, there 
are visible deviations in the tails: the histogram shows slightly heavier tails 
than the Gaussian curve, particularly on the left side. 


## c.

```{r}
kde_bcv <- density(ba_log_returns, bw = "bcv")
mu_mle <- mean(ba_log_returns)
sigma_mle <- sd(ba_log_returns)
gaussian_y <- dnorm(kde_bcv$x, mean = mu_mle, sd = sigma_mle)

plot_df <- data.frame(
  x = kde_bcv$x,
  KDE = kde_bcv$y,
  Gaussian = gaussian_y
)


library(ggplot2)
ggplot(plot_df) +
  geom_line(aes(x = x, y = KDE, color = "KDE"), linewidth = 1.2) +
  geom_line(aes(x = x, y = Gaussian, color = "Gaussian"), linetype = "dashed", linewidth = 1.2) +
  scale_color_manual(name = "Density Type", values = c("KDE" = "blue", "Gaussian" = "red")) +
  labs(
    title = "KDE vs. Gaussian Fit for Boeing (BA) Log Returns (BCV)",
    x = "Log Return", y = "Density"
  ) +
  theme_minimal()



```
Guassian is underestimating the peak as well at the tails compared to KDE. 

## d.
```{r}

library(ggplot2)
library(reshape2)

stock_names <- colnames(log_returns)[colnames(log_returns) != "Date"]

density_list <- list()

for (stock in stock_names) {
  stock_returns <- log_returns[[stock]]
  kde_result <- density(stock_returns, bw = "bcv")
  temp_df <- data.frame(
    x = kde_result$x,
    density = kde_result$y,
    Stock = stock
  )
  density_list[[stock]] <- temp_df
}
all_kde_data <- do.call(rbind, density_list)
ggplot(all_kde_data, aes(x = x, y = density, color = Stock)) +
  geom_line(alpha = 0.8) +
  labs(
    title = "Kernel Density Estimates of Log Returns for 28 Stocks",
    x = "Log Return", y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "none") 

```


The other curves for the other 27 stocks look similar to the BA curves, and they 
seem to support my previous answer that Guassian is underestimating the peak as 
well at the tails. 

\newpage


## Problem 3

## a.

```{r}
log_returns_matrix <- diff(log(as.matrix(close_price)))
factor_model <- factanal(log_returns_matrix, factors = 1, scores = "regression")
factor_loadings <- factor_model$loadings[, 1]
sorted_loadings <- sort(factor_loadings)

barplot(sorted_loadings,
        las = 2,
        main = "Factor Loadings from One-Factor Model",
        xlab = "Stocks",
        ylab = "Factor Loadings")



```

\newpage

```{r}
library(knitr)

factor_loading_table <- data.frame(
  Stock = names(factor_loadings),
  Loading = factor_loadings
)

factor_loading_table <- factor_loading_table[order(factor_loading_table$Loading), ]

kable(factor_loading_table,
      caption = "Factor Loadings from One-Factor Model",
      digits = 4,
      align = c("l", "r"))


```




## b.

The one-factor model appears to extract a market-wide risk factor, possibly 
related to macroeconomic or financial conditions. Stocks in sectors like financials
and industrials load heavily(e.x. JPM, GS), while more defensive or sector-specific 
firms (like DD, WMT, AXP) load lower. This distribution aligns with economic 
intuition: some companies are inherently more cyclically exposed than others.

## c.


```{r}
factor_scores <- factor_model$scores[, 1]
score_dates <- as.Date(rownames(close_price))[-1]

plot(score_dates, factor_scores, type = "l",
     col = "darkblue",
     xlab = "Date", ylab = "Factor Score",
     main = "Factor Scores from One-Factor Model")

  
```

The factor scores fluctuate around zero, as expected in a factor model.

There are some noticeable spikes, indicating market-wide shocks or volatility clusters.

One major drop and sharp recovery occurs around late August to early September 2015. 

### d.

The overall trend is similar with spikes and an obvious drop around late August to
early September. 
The factor score plot, however, distills the co-movement across all 
stocks into a single series. It’s much smoother and captures shared market behavior 
rather than individual noise.

\newpage

## Q4.

```{r}
library(boot)
library(ggplot2)


resample.cases <- function() {
  resampled_data <- log_returns[sample(1:nrow(log_returns), replace = TRUE), ]
  return(resampled_data)
}

factor_loading_estimator <- function(data) {
  fa_model <- factanal(data, factors = 1, scores = "regression")
  return(fa_model$loadings[, 1])
}

set.seed(402)
B <- 400
bootstrap_loadings <- replicate(B, factor_loading_estimator(resample.cases()))

equitails <- function(x, alpha) {
  lower <- quantile(x, alpha / 2)
  upper <- quantile(x, 1 - alpha / 2)
  return(c(lower, upper))
}

ci_85 <- apply(bootstrap_loadings, 1, equitails, alpha = 0.15)

fa_original <- factanal(log_returns, factors = 1, scores = "regression")
original_loadings <- fa_original$loadings[, 1]

ci_df <- data.frame(
  Stock = rownames(fa_original$loadings),
  Estimate = original_loadings,
  Lower_85_CI = ci_85[1, ],
  Upper_85_CI = ci_85[2, ]
)

ggplot(ci_df, aes(x = reorder(Stock, Estimate), y = Estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = Lower_85_CI, ymax = Upper_85_CI), width = 0.2) +
  theme_minimal() +
  coord_flip()+
  labs(
    title = "Bootstrapped 85% CI for Factor Loadings",
    x = "Stocks",
    y = "Factor Loadings"
  )


```

\newpage

### Q5. 
```{r}
cov_GE_CVX <- cov(log_returns$GE, log_returns$CVX) 
cov_GE_BA <- cov(log_returns$GE, log_returns$BA)    
cov_GS_CVX <- cov(log_returns$GS, log_returns$CVX)  
cov_GS_BA <- cov(log_returns$GS, log_returns$BA)  


```

Sample Covariances:
GE and Chevron:  $1.23 \times 10^{-4}$
GE and Boeing: $9.32 \times 10^{-5}$
Goldman Sachs and Chevron: $1.34 \times 10^{-4}$
Goldman Sachs and Boeing: $1.14 \times 10^{-4}$

## b.

```{r}

factor_loadings <- factor_model$loadings[, 1]  
factor_scores <- factor_model$scores[, 1]
sd_GE <- sd(log_returns$GE)
sd_CVX<-sd(log_returns$CVX)
sd_BA<-sd(log_returns$BA)
sd_GS<-sd(log_returns$GS)
pred_cov_GE_CVX <- factor_loadings["GE"] * factor_loadings["CVX"]*sd_GE*sd_CVX
pred_cov_GE_BA <- factor_loadings["GE"] * factor_loadings["BA"]*sd_GE*sd_BA
pred_cov_GS_CVX <- factor_loadings["GS"] * factor_loadings["CVX"]*sd_CVX*sd_GS
pred_cov_GS_BA <- factor_loadings["GS"] * factor_loadings["BA"]*sd_GS*sd_BA


```
The covariance of GE and Chevron is $1.10 \times 10^{-4}$

GE and Boeing is  $9.79 \times 10^{-5}$

Goldman Sachs and Chevron is  $1.32 \times 10^{-4}$

Goldman Sachs and Boeing: $1.17 \times 10^{-4}$



### c.



The one-factor model implies that the covariance between two assets is:

\[
\text{Cov}[i, j] = w_i w_j \cdot sd_i \cdot sd_j
\]

where \( w_i \) and \( w_j \) are the factor loadings for assets \( i \) and
\( j \), and \( sd_i \), \( sd_j \) are their sample standard deviations.

Applying this to our four covariances:

\[
\text{Cov}[GE, CVX] = w_{GE} w_{CVX} \cdot sd_{GE} \cdot sd_{CVX}
\]
\[
\text{Cov}[GE, BA] = w_{GE} w_{BA} \cdot sd_{GE} \cdot sd_{BA}
\]
\[
\text{Cov}[GS, CVX] = w_{GS} w_{CVX} \cdot sd_{GS} \cdot sd_{CVX}
\]
\[
\text{Cov}[GS, BA] = w_{GS} w_{BA} \cdot sd_{GS} \cdot sd_{BA}
\]

Now compute the full ratio:

\[
\frac{\frac{\text{Cov}[GE, CVX]}{\text{Cov}[GE, BA]}}{\frac{\text{Cov}[GS, CVX]}{\text{Cov}[GS, BA]}} 
= \frac{\frac{w_{GE} w_{CVX} \cdot sd_{GE} sd_{CVX}}{w_{GE} w_{BA} \cdot sd_{GE} sd_{BA}}}
       {\frac{w_{GS} w_{CVX} \cdot sd_{GS} sd_{CVX}}{w_{GS} w_{BA} \cdot sd_{GS} sd_{BA}}}
\]


\[
= \frac{\frac{w_{CVX} \cdot sd_{CVX}}{w_{BA} \cdot sd_{BA}}}
        {\frac{w_{CVX} \cdot sd_{CVX}}{w_{BA} \cdot sd_{BA}}} = 1
\]

So this compound ratio of covariances should equal 1, since all the stock standard 
deviations and factor terms cancel out.


## d.

```{r}
set.seed(402)
library(boot)

resample.cases <- function() {
  log_returns[sample(1:nrow(log_returns), replace = TRUE), ]
}

cov_ratio_estimator <- function(data) {
  num <- cov(data[, "GE"], data[, "CVX"]) / cov(data[, "GE"], data[, "BA"])
  denom <- cov(data[, "GS"], data[, "CVX"]) / cov(data[, "GS"], data[, "BA"])
  return(num / denom)
}

equitails <- function(x, alpha) {
  lower <- quantile(x, alpha / 2)
  upper <- quantile(x, 1 - alpha / 2)
  return(c(lower, upper))
}

rboot <- function(statistic, simulator, B) {
  replicate(B, statistic(simulator()))
}

bootstrap.ci <- function(statistic, simulator, B, t.hat, level = 0.95) {
  tboots <- rboot(statistic, simulator, B)
  alpha <- 1 - level
  ci <- equitails(tboots, alpha)
  lower <- t.hat + (t.hat - ci[2])
  upper <- t.hat + (t.hat - ci[1])
  return(cbind(lower = lower, upper = upper))
}

set.seed(402)
t.hat <- cov_ratio_estimator(log_returns)

cov_ratio_ci <- bootstrap.ci(
  statistic = cov_ratio_estimator,
  simulator = resample.cases,
  B = 400,
  t.hat = t.hat,
  level = 0.95
)





```


The 95% bootstrapping CI is  [ 0.8727 , 1.3232 ] which contains 1, so the observed 
covariances are consistent with the one-factor model.

\newpage

### Q6


```{r}
library(dplyr)
f2_model <- factanal(log_returns, factors = 2, scores = "regression")

loadings_df <- as.data.frame(f2_model$loadings[, 1:2])

loadings_df$Asset <- rownames(loadings_df)

loadings_df_f1 <- loadings_df %>%
  arrange(Factor1) %>%
  mutate(Asset = factor(Asset, levels = Asset))

ggplot(loadings_df_f1, aes(x = Asset, y = Factor1)) +
  geom_bar(stat = "identity") +
  labs(title = "Factor 1 Loadings", x = "Assets", y = "Loading") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

loadings_df_f2 <- loadings_df %>%
  arrange(Factor2) %>%
  mutate(Asset = factor(Asset, levels = Asset))

ggplot(loadings_df_f2, aes(x = Asset, y = Factor2)) +
  geom_bar(stat = "identity") +
  labs(title = "Factor 2 Loadings", x = "Assets", y = "Loading") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



library(knitr)

loadings_table <- loadings_df %>%
  select(Asset, Factor1, Factor2) %>%
  arrange(Factor1)

kable(loadings_table, 
      caption = "Factor Loadings from Two-Factor Model",
      digits = 4,
      align = c("l", "r", "r"))



```

For the two-factor model, some assets that had very high Factor 1 loadings in the 
one-factor model now  have reduced loadings.This may suggests that these assets 
are influenced by an additional, more specific risk factor captured by Factor 2. 
Some assets like CVX, XOM, CAT now have much lower loadings compared to others, 
suggesting the factor is no longer just capturing general market exposure. As a result, 
Factor 1 has shifted from a pure market-wide factor to one that may now capture a 
more sector-specific or thematic variation, due to the introduction of Factor 2, 
which absorbs part of the general market co-movement.




\newpage

## Q7. 
```{r}
log_tricky <- diff(log(as.matrix(tricky_prices)))
score_dates <- as.Date(rownames(tricky_prices))[-1]
f1_tricky <- factanal(log_tricky, factors = 1, scores = "regression")
factor_loadings <- f1_tricky$loadings[, 1]
sorted_loadings <- sort(factor_loadings)
barplot(sorted_loadings,
        las = 2,
        main = "Factor Loadings from One-Factor Model",
        xlab = "Stocks",
        ylab = "Factor Loadings")
factor_scores <- f1_tricky$scores[, 1]



library(knitr)

factor_loading_table <- data.frame(
  Stock = names(factor_loadings),
  Loading = factor_loadings
)

factor_loading_table <- factor_loading_table[order(factor_loading_table$Loading), ]


kable(factor_loading_table, 
      caption = "Factor Loadings from One-Factor Model",
      digits = 3, 
      align = c("l", "r"))
```

```{r}
plot(score_dates, factor_scores, type = "l",
     col = "darkblue",
     xlab = "Date", ylab = "Factor Score",
     main = "Factor Scores from One-Factor Model (30 Stocks)")

```

```{r}
plot(score_dates, log_tricky[, 1], type = "n",
     xlab = "Date", ylab = "Log Daily Return",
     main = "Log Daily Returns of 30 US Stocks")
colors <- rainbow(ncol(log_tricky))
for (i in 1:ncol(log_tricky)) {
  lines(score_dates, log_tricky[, i], col = adjustcolor(colors[i], alpha.f = 0.5))
}
```


When the two new stocks, Visa (V) and Nike (NKE), were added to the dataset, there 
was a noticeable change in the distribution of factor loadings. In the original 
plot with 28 large US corporations, all stocks had relatively high factor loadings, 
clustering around 0.5 to 0.8, indicating a strong relationship with the dominant 
market factor. However, after adding V and NKE, both appeared on the far left of the 
bar chart, showing significantly lower factor loadings. This suggests that these 
two companies are less correlated with thecommon market factor driving the other
28 stocks. As a result, the overall variation in factor exposures increased, 
and the previously tight cluster of high factor loadings became more spread out.

Comparing the factor scores from the one-factor model before and after 
adding Visa (V) and Nike (NKE), we see that the overall pattern remains similar, 
with scores fluctuating around zero and capturing market-wide shocks like the volatility
spike in late August 2015,  indicating that the core structure of the market-driven 
variation is preserved.

As the factor scores, the overall pattern of log daily returns remained largely 
similar before and after adding the 2 stocks. This suggests that the newly added 
stocks behaved similarly to the original group or were not volatile enough to noticeably 
alter the aggregate pattern.

\newpage

## Q8.

7 hours